Understanding Parameterized Quantum Circuit Learning for Quantum Chemical Applications

### 2025-04-09 manuscript/main.tex
Wed  9 Apr 2025 17:01:28 EDT

While most QC studies that are relevant to computational chemists focus on creating more efficient electronic structure methods on quantum computers\cite{romero_strategies_2019,mcardle_quantum_2020,bauer_quantum_2020,cao_quantum_2019}, an approach that combines both ML and QC, is quantum machine learning (QML).
Using either formal mathematical proofs or numerical results based on empirical observations, QML has shown potential quantum speedups for various applications using a diverse set of implementations.\cite{biamonte_quantum_2017}
While several classes of QML algorithms have shown promise for providing flexible ML models, parameterized quantum circuits (PQCs) are capable of achieving non-trivial results on near-term quantum hardware.
PQCs formulate the ML algorithm as a variational problem optimized using a hybrid approach using both classical and quantum hardware.\cite{benedetti_parameterized_2019}
Like classical ML approaches, PQCs have been applied for several chemistry use cases such as drug\cite{suzuki_predicting_2020,smaldone_quantum--classical_2024,bhatia_quantum_2023,kao_exploring_2023,li_quantum_2021,avramouli_quantum_2023,avramouli_unlocking_2023} and materials discovery\cite{ishiyama_noise-robust_2022,ryu_quantum_2023,vitz_hybrid_2024}, the prediction of proton affinities\cite{jin_integrating_2025}, and experimental molecular properties, including the log solubility in water, melting point, octanol/water distribution coefficient, and hydration free energy of small molecules in water.\cite{hatakeyama-sato_quantum_2023}
Despite the broad range of topics covered in these studies and the interest among computational chemists in exploring PQCs for chemical applications, studies analyzing the potential benefits or drawbacks of using QML for quantum chemistry are lacking.

@@

To maintain a general notation, in the following section, we follow the notations derived from  Ref. \citep{suzuki_predicting_2020}.
Encoding layers work by mapping a given $d$-dimensional feature vector, $\mathbf{x}=(x_{1}, x_{2}, \ldots, x_{d})^{T} \in \mathbb{R}^{d}$, normalized on the range $[-1,1]$, onto a quantum circuit using a unitary matrix, denoted as $U_{\Phi(\mathbf{x})}$, to produce the quantum state $U_{\Phi(\mathbf{x})}\ket{0}^{\otimes n}$, where $n$ is the number of qubits.
More generally, the encoding layer can also incorporate both unitary and entanglement gates, such that 
\begin{equation}
	U_{\Phi(x)} =  \prod_{l} E_{\text{ent}}^{l} U_{\phi_{l}(\mathbf{x})},
	\label{eq:general_encoding}
\end{equation}
where, $E_{\text{ent}}^{l}$ denotes the entangling gates, e.g., CNOT or CZ, and  $U_{\phi_{l}(\mathbf{x})}$ denotes the choice of encoding unitaries.

In this study, like in Ref. \citep{suzuki_predicting_2020}, we explore encoders with $l={1,2}$. 
When $l=1$, this forms the simplest encoding layers with $E_{\text{ent}}^{1}$ being absent, e.g. corresponding to the indenty ($\mathbf{I}$). 
Using the previously defined general notation, the simplest encoding layer, single-angle encoding (Fig. \ref{fig:encoders} \textbf{(a)}), is then defined as,
\begin{equation}
	U_{\text{A1}} = \prod_{i=0}^{n} R^{Y}_{i}(x_{i}),
	\label{eq:A1}
\end{equation}
where $R^{Y}_{i}$ denotes a parameterized Y rotation gate on qubit $i$.
Single-angle encoding, or qubit encoding, is structured similarly to a product of unentangled quantum states and have a similar mathematical structure as a product of local kernels where each $x_{i}$ is encoded in a local feature map.\cite{stoudenmire_supervised_2016,benedetti_parameterized_2019}

@@

When $l=2$, more complex circuits can be generated, like in Ref. \citep{suzuki_predicting_2020}, where we choose the entanglement gates, $E_{\text{ent}}^{1}$ and $E_{\text{ent}}^{2}$, to be equivalent and the encoding layer takes the following form, $U_{\Phi(x)} =  E_{\text{ent}} U_{\phi_{2}(\mathbf{x})} E_{\text{ent}} U_{\phi_{1}(\mathbf{x})}$ .
Additionally, we exclude IQP encoding when $l=2$ due to the increased computational costs associated with the circuit depth when compared to the expanded models based on A1, A2, and M encoding.
Therefore, when we exclude IQP, there are five unique combinations of $U_{\phi_{1}(\mathbf{x})}$ and $U_{\phi_{2}(\mathbf{x})}$ (M-M, A1-A1, A2-A2, M-A1, and M-A2) and two different entanglement layer options (CNOT and CZ) for a total of 10 encoding circuits with $l=2$.
The common notation for these circuits is $U_{\phi_{1}(\mathbf{x})}-U_{\phi_{2}(\mathbf{x})}-E_{\text{ent}}$ where two example encoding circuits are M--M--CNOT and M--A1--CNOT.
The 14 encoding circuits used throughout this study are shown in Table \ref{tab:encoders}, where the first column shows the label, second column the first unitary ($U_{\phi_{1}(\mathbf{x})}$), third column the second unitary ($U_{\phi_{2}(\mathbf{x})}$), and last column the entanglement gates ($E_{\text{ent}}$).

\begin{table}[htbp]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		\textbf{Name} & $U_{\phi_{1}(\mathbf{x})}$ & $U_{\phi_{2}(\mathbf{x})}$ & $E_{\text{ent}}$  \\
		\hline
		\hline
		A1 & $U_{\text{A1}}$ & --- & --- \\
		\hline
		A2 & $U_{\text{A2}}$ & --- & --- \\
		\hline		
		M & $U_{\text{M}}$ & --- & --- \\
		\hline
		IQP & $U_{\text{IQP}}$ & --- & --- \\
		\hline
		A1--A1--CNOT & $U_{\text{A1}}$ & $U_{\text{A1}}$ & $E_{\text{CNOT}}$ \\
		\hline
		 A2--A2--CNOT & $U_{\text{A2}}$ & $U_{\text{A2}}$ & $E_{\text{CNOT}}$ \\
		\hline
		M--M--CNOT & $U_{\text{M}}$ & $U_{\text{M}}$ & $E_{\text{CNOT}}$ \\
		\hline
		M--A1--CNOT & $U_{\text{M}}$ & $U_{\text{A1}}$ & $E_{\text{CNOT}}$ \\
		\hline		
		M--A2--CNOT & $U_{\text{M}}$ & $U_{\text{A2}}$ & $E_{\text{CNOT}}$ \\
		\hline				
		A1--A1--CZ & $U_{\text{A1}}$ & $U_{\text{A1}}$ & $E_{\text{CZ}}$ \\
		\hline
		A2--A2--CZ& $U_{\text{A2}}$ & $U_{\text{A2}}$ & $E_{\text{CZ}}$ \\
		\hline
		M--M--CZ & $U_{\text{M}}$ & $U_{\text{M}}$ & $E_{\text{CZ}}$ \\
		\hline
		M--A1--CZ & $U_{\text{M}}$ & $U_{\text{A1}}$ & $E_{\text{CZ}}$ \\
		\hline		
		M--A2--CZ & $U_{\text{M}}$ & $U_{\text{A2}}$ & $E_{\text{CZ}}$ \\
		\hline						
	\end{tabular}
	\caption{The 14 encoding layers used in this study where the first column shows the encoding circuit name, second column the unitary for $l=1$, third column the unitary for $l=2$, and last column the corresponding entanglement gates.}
	\label{tab:encoders}
\end{table}

@@

The last component of a PQC is measurement, which is required to recover the predicted target values, $\hat{y}_{i}$, of the machine learning model that is used to update the model parameters.
This is performed by measuring the quantum state, $\ket{\Psi}$, using the Pauli Z operator on the first qubit denoted as,
\begin{equation}
	\hat{y}_{i} = \bra{\Psi}Z_{0}\ket{\Psi}_{i},
	\label{eq:y_pred}
\end{equation}
and passing the set of predicted target values, $\bm{\hat{y}} = (\hat{y}_{1}, \ldots, \hat{y}_{N}) \in \mathbb{R}^{N}$, where $N$ is the number of samples, to the loss function, $\mathcal{L}(\bm{y}, \bm{\hat y})$, where $y_{i}$ belongs to the set of true target values $\bm{y} = (y_{1}, \ldots, y_{N}) \in \mathbb{R}^{N}$.
While $\mathcal{L}$ can be any loss function relevant for regression-based ML tasks, we choose to mean square error to be the loss function,
\begin{equation}
	\mathcal{L}(\bm{y}, \bm{\hat y}) = \frac{1}{N} \sum_{i=1}^{N} (y_{i} - \hat{y}_{i})^{2}.
	\label{eq:isthisloss}
\end{equation}

\section{Implementation}
To explore the large combination of PQCs previously mentioned, we introduce \textit{qregress} (Fig. \ref{fig:overview_project}), a modular Python package based on PennyLane\cite{bergholm_pennylane_2022} and Qiskit\cite{javadi-abhari_quantum_2024}.
The portion of the code based on PennyLane is capable of performing state-vector simulation using Qulacs\cite{suzuki_qulacs_2021}, noisy simulation using \textit{qiskit-aer} with the \textit{FakeQuebec} backend, along with access to real devices accessible on the IBM Quantum Platform using the PennyLane-Qiskit plugin.
Additionally, to train and test regression models that require many circuit executions, we ported our PennyLane code to Qiskit so that the Qiskit Batch Execution mode could be utilized.
This was a vital step for running on \textit{ibm\_quebec} since Qiskit Sessions, as implemented in the PennyLane-Qiskit plugin, are too cumbersome to train PQCs as $N$, the number of training samples, increases.
In practice, all state-vector simulations were performed using PennyLane, while all calculations using \textit{FakeQuebec} and \textit{ibm\_quebec}  are run using Qiskit.
All state-vector simulations with 5 qubits were optimized using 1000 iterations, while 16 qubit are initially ran using 250 iterations and longer runs, using 1000 iterations, use early stopping due to computational cost considerations.
State-vector simulations were also optimized using the Simultaneous Perturbation Stochastic Approximation method (SPSA), as implemented in PennyLane since we found that SPSA requires less steps to optimize when compared to other optimizers for state-vector simulation. 
For the experiments ran using \textit{FakeQuebec} and  \textit{ibm\_quebec}, we utilized the Constrained Optimization By Linear Approximation (COBYLA) optimizer, as implemented in SciPy\cite{virtanen_scipy_2020}.
For all models, features ($\mathbf{x}$) and target values ($\mathbf{y}$) were scaled using the MinMaxScaler, as implemented in Scikit-learn\cite{pedregosa_scikit-learn_2011}, such that all features and target values are $\mathbb{R}\in [ -1,1 ]$.

@@

An important aspect of ML in chemistry is the choice of molecular representation, or how the molecule is represented in the machine learning models.\cite{jones_molecular_2023}
In general, molecular representations can be partitioned into three groups, graph-, topology-, and physics-based representations, all of which were explored to provide a comprehensive overview of how they perform using classical models as a baseline.
During the preprocessing stage, before converting to the XYZ coordinates into our molecular representation of choice, the set of hypothetical structures was reduced to 2436 molecules due to issues with valency exceptions when converting XYZ coordinates into RDKit mol objects.
Using RDKit\cite{noauthor_rdkit_nodate} we examined three commonly applied graph-based molecular preresentations, Molecular ACCess Systems (MACCS)\cite{durant_reoptimization_2002}, Morgan or extended-connectivity fingerprints \cite{morgan_generation_1965,rogers_extended-connectivity_2010}, and RDKit fingerprints.
All three of these methods use traversals of the molecular graphs to encode various structural details into bit vectors.
Lastly, we explore both topology- and physics-based molecular representations, both of which encode the three-dimensional structure of molecules in various, unique ways.
Persistent images (PIs) are a topology-based fingerprint that uses persistence homology to encode topological information of three-dimensional molecular structures into fixed-dimension images.\cite{adams_persistence_2017,townsend_representation_2020,schiff_augmenting_2022} 
We use the implementation from Townsend \textit{et al.}\cite{townsend_representation_2020}, which uses the Ripser Python package to generate PIs.\cite{tralie_ripserpy_2018}
Lastly, we explore two physics-based representations, Coulomb matrices (CMs) \cite{rupp_fast_2012} and smooth overlap of atomic positions (SOAPs), that were generated using DScribe.\cite{de_comparing_2016}
Both methods encode physical information regarding the atomic environments of each molecule, where CMs encode the Coulomb repulsion between atoms using nuclear charges, while SOAPs encode the similarities between atomic neighborhoods using kernels.
Due to the computational cost associated with computing the regularized entropy match (REMatch) kernel over the set of 2436 molecules with the SOAPs representation, we excluded SOAPs from the overall discussion.
For each of the molecular representations, we analyzed two different methods for representing the components of the BSE reaction.
The first represents the feature vector as the products subtracted from the reactants, denoted as \textit{sub} (e.g., $\mathbf{X}_{\mathit{sub}} = (\mathbf{X}_{\text{A}^{.}} + \mathbf{X}_{\text{B}^{.}}) - \mathbf{X}_{\text{A-B}}$), similar to the method used in Ref. \cite{garcia-andrade_barrier_2023}, and the second represents the feature vector using only the reactant (A-B), denoted as \textit{AB} (e.g., $\mathbf{X}_{\mathit{AB}} =  \mathbf{X}_{\text{A-B}}$).

Using this set of molecular representations, we also examine a diverse set of classical regression models to help choose the optimal representation for our diverse set of PQCs.
The classical models we analyze include ridge, lasso, elastic net, \textit{k}-nearest-neighbors, random forest, gradient boosting, support vector machines, kernel ridge, and Gaussian process regression as implemented in sckit-learn.\cite{pedregosa_scikit-learn_2011}
As highlighted in Fig. \ref{fig:classical_molrepfig}, we found that the best molecular representation across all models test was Morgan fingerprints using the \textit{sub} formulation, which we use for all experiments using the PQCs.
Additionally, one challenge of applying classical molecular representations for ML tasks using PQCs is reducing the set of features, often containing hundreds or thousands of features per sample, to the number of qubits used on the quantum device.
For example, at the onset of this project the standard number of qubits on the IBM quantum devices was 5 or 16 qubits, which requires reducing the initial set of 2048 features per sample for Morgan fingerprints down to 5 or 16 features.
To perform this reduction, we explored two different methods, SHapley Additive ExPlanation analysis (SHAP)\cite{lundberg_unified_2017}, which uses cooperative game theory to determine feature importance, and principal component analysis (PCA), which is a dimensionality reduction technique that captures the direction a datasets variance, as implemented in scikit-learn.\cite{pedregosa_scikit-learn_2011}
In Figs. \ref{fig:BSE_classical_features_R2} and \ref{fig:BSE_bse_classical_features_MAE}, we compare the two reduction techniques, for both 5 and 16 features, with the initial model containing 2048 features.
The initial model has mean absolute errors (MAE) of 1.91 and 4.98 kcal/mol and R$^{2}$s of 0.99 and 0.91 for the training and test set, respectively.
Using SHAP, we found that the model containing 5 features has an MAE of 16.08 kcal/mol and an R$^{2}$ of 0.39 for the training set, while the test set has an MAE of 15.86 kcal/mol and an R$^{2}$ of 0.42.
For the model with 16 features, we observe improvements in both the training and test sets, with MAEs of 10.48 and 11.08 kcal/mol and R$^{2}$ values of 0.69 and 0.68, respectively.
For the models using 5 and 16 features reduced through PCA, the training sets yield MAEs of 4.09 and 3.23 kcal/mol, while the test sets have MAEs of 10.17 and 8.40 kcal/mol, respectively.
The R$^{2}$s also show slight improvement over the feature reductions using SHAP, where the R$^{2}$s using 5 features are 0.95 and 0.69 for the training and test set, respectively, and the model using 16 features has R$^{2}$s of 0.97 and 0.78 for the training and test set, respectively.
Overall, the models using SHAP to reduce the feature set offer models with less overfitting, while sacrificing accuracy, when compared to the models reduced using PCA, which offer more accurate models that are more overfit. 
Based on these insights, for the experiments using PQCs, all BSE49 data is reduced to 5 or 16 qubits using PCA applied on Morgan fingerprint using the \textit{sub} formulation.

>>>>

While most QC studies that are relevant to computational chemists focus on creating more efficient electronic structure methods on quantum computers\cite{romero_strategies_2019,mcardle_quantum_2020,bauer_quantum_2020,cao_quantum_2019}, an approach that combines both ML and QC, is quantum machine learning (QML).
Using either formal mathematical proofs or numerical results based on empirical observations, QML has shown potential quantum speedups for various applications using a diverse set of implementations.\cite{biamonte_quantum_2017}
While several classes of QML algorithms have shown promise for providing flexible ML models, parameterized quantum circuits (PQCs) can achieve non-trivial results on near-term quantum hardware.
PQCs formulate the ML algorithm as a variational problem optimized using a hybrid approach using classical and quantum hardware.\cite{benedetti_parameterized_2019}
Like classical ML approaches, PQCs have been applied for several chemistry use cases such as drug\cite{suzuki_predicting_2020,smaldone_quantum--classical_2024,bhatia_quantum_2023,kao_exploring_2023,li_quantum_2021,avramouli_quantum_2023,avramouli_unlocking_2023} and materials discovery\cite{ishiyama_noise-robust_2022,ryu_quantum_2023,vitz_hybrid_2024}, the prediction of proton affinities\cite{jin_integrating_2025}, and experimental molecular properties, including the log solubility in water, melting point, octanol/water distribution coefficient, and hydration free energy of small molecules in water.\cite{hatakeyama-sato_quantum_2023}
Despite the broad range of topics covered in these studies and the interest among computational chemists in exploring PQCs for chemical applications, studies analyzing the potential benefits or drawbacks of using QML for quantum chemistry are lacking.

@@

In the following section, we adopt the notations derived from Ref. \citep{suzuki_predicting_2020} to maintain consistency.
Encoding layers work by mapping a given $d$-dimensional feature vector, $\mathbf{x}=(x_{1}, x_{2}, \ldots, x_{d})^{T} \in \mathbb{R}^{d}$, normalized on the range $[-1,1]$, onto a quantum circuit using a unitary matrix, denoted as $U_{\Phi(\mathbf{x})}$, to produce the quantum state $U_{\Phi(\mathbf{x})}\ket{0}^{\otimes n}$, where $n$ is the number of qubits.
More generally, the encoding layer can also incorporate both unitary and entanglement gates, such that 
\begin{equation}
	U_{\Phi(x)} =  \prod_{l} E_{\text{ent}}^{l} U_{\phi_{l}(\mathbf{x})},
	\label{eq:general_encoding}
\end{equation}
where, $E_{\text{ent}}^{l}$ denotes the entangling gates, e.g., CNOT or CZ, and  $U_{\phi_{l}(\mathbf{x})}$ denotes the choice of encoding unitaries.

In this study, like in Ref. \citep{suzuki_predicting_2020}, we explore encoders with $l={1,2}$. 
When $l=1$, this forms the simplest encoding layers with $E_{\text{ent}}^{1}$ being absent, e.g. corresponding to the indenty ($\mathbf{I}$). 
Using the previously defined general notation, the simplest encoding layer, single-angle encoding (Fig. \ref{fig:encoders} \textbf{(a)}), is then defined as,
\begin{equation}
	U_{\text{A1}} = \prod_{i=0}^{n} R^{Y}_{i}(x_{i}),
	\label{eq:A1}
\end{equation}
where $R^{Y}_{i}$ denotes a parameterized Y rotation gate on qubit $i$.
Single-angle encoding, or qubit encoding, is structured similarly to a product of unentangled quantum states and has a similar mathematical structure as a product of local kernels where each $x_{i}$ is encoded in a local feature map.\cite{stoudenmire_supervised_2016,benedetti_parameterized_2019}

@@

When $l=2$, more complex circuits can be generated, like in Ref. \citep{suzuki_predicting_2020}, where we choose the entanglement gates, $E_{\text{ent}}^{1}$ and $E_{\text{ent}}^{2}$, to be equivalent and the encoding layer takes the following form, $U_{\Phi(x)} =  E_{\text{ent}} U_{\phi_{2}(\mathbf{x})} E_{\text{ent}} U_{\phi_{1}(\mathbf{x})}$.
Additionally, we exclude IQP encoding when $l=2$ due to the increased computational costs associated with the circuit depth when compared to the expanded models based on A1, A2, and M encoding.
Therefore, when we exclude IQP, there are five unique combinations of $U_{\phi_{1}(\mathbf{x})}$ and $U_{\phi_{2}(\mathbf{x})}$ (M-M, A1-A1, A2-A2, M-A1, and M-A2) and two different entanglement layer options (CNOT and CZ) for a total of 10 encoding circuits with $l=2$.
The common notation for these circuits is $U_{\phi_{1}(\mathbf{x})}-U_{\phi_{2}(\mathbf{x})}-E_{\text{ent}}$ where two example encoding circuits are M--M--CNOT and M--A1--CNOT.
The 14 encoding circuits used throughout this study are shown in Table \ref{tab:encoders}, where the first column shows the label, second column the first unitary ($U_{\phi_{1}(\mathbf{x})}$), third column the second unitary ($U_{\phi_{2}(\mathbf{x})}$), and last column the entanglement gates ($E_{\text{ent}}$).

\begin{table}[htbp]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		\textbf{Name} & $U_{\phi_{1}(\mathbf{x})}$ & $U_{\phi_{2}(\mathbf{x})}$ & $E_{\text{ent}}$  \\
		\hline
		\hline
		A1 & $U_{\text{A1}}$ & --- & --- \\
		\hline
		A2 & $U_{\text{A2}}$ & --- & --- \\
		\hline		
		M & $U_{\text{M}}$ & --- & --- \\
		\hline
		IQP & $U_{\text{IQP}}$ & --- & --- \\
		\hline
		A1--A1--CNOT & $U_{\text{A1}}$ & $U_{\text{A1}}$ & $E_{\text{CNOT}}$ \\
		\hline
		 A2--A2--CNOT & $U_{\text{A2}}$ & $U_{\text{A2}}$ & $E_{\text{CNOT}}$ \\
		\hline
		M--M--CNOT & $U_{\text{M}}$ & $U_{\text{M}}$ & $E_{\text{CNOT}}$ \\
		\hline
		M--A1--CNOT & $U_{\text{M}}$ & $U_{\text{A1}}$ & $E_{\text{CNOT}}$ \\
		\hline		
		M--A2--CNOT & $U_{\text{M}}$ & $U_{\text{A2}}$ & $E_{\text{CNOT}}$ \\
		\hline				
		A1--A1--CZ & $U_{\text{A1}}$ & $U_{\text{A1}}$ & $E_{\text{CZ}}$ \\
		\hline
		A2--A2--CZ& $U_{\text{A2}}$ & $U_{\text{A2}}$ & $E_{\text{CZ}}$ \\
		\hline
		M--M--CZ & $U_{\text{M}}$ & $U_{\text{M}}$ & $E_{\text{CZ}}$ \\
		\hline
		M--A1--CZ & $U_{\text{M}}$ & $U_{\text{A1}}$ & $E_{\text{CZ}}$ \\
		\hline		
		M--A2--CZ & $U_{\text{M}}$ & $U_{\text{A2}}$ & $E_{\text{CZ}}$ \\
		\hline						
	\end{tabular}
	\caption{The 14 encoding layers used in this study where the first column shows the encoding circuit name, the second column the unitary for $l=1$, the third column the unitary for $l=2$, and the last column the corresponding entanglement gates.}
	\label{tab:encoders}
\end{table}

@@

The last component of a PQC is the measurement, which is required to recover the predicted target values, $\hat{y}_{i}$, of the machine learning model that is used to update the model parameters.
This is performed by measuring the quantum state, $\ket{\Psi}$, using the Pauli Z operator on the first qubit denoted as,
\begin{equation}
	\hat{y}_{i} = \bra{\Psi}Z_{0}\ket{\Psi}_{i},
	\label{eq:y_pred}
\end{equation}
and passing the set of predicted target values, $\bm{\hat{y}} = (\hat{y}_{1}, \ldots, \hat{y}_{N}) \in \mathbb{R}^{N}$, where $N$ is the number of samples, to the loss function, $\mathcal{L}(\bm{y}, \bm{\hat y})$, where $y_{i}$ belongs to the set of true target values $\bm{y} = (y_{1}, \ldots, y_{N}) \in \mathbb{R}^{N}$.
While $\mathcal{L}$ can be any loss function relevant for regression-based ML tasks, we choose to mean square error to be the loss function,
\begin{equation}
	\mathcal{L}(\bm{y}, \bm{\hat y}) = \frac{1}{N} \sum_{i=1}^{N} (y_{i} - \hat{y}_{i})^{2}.
	\label{eq:isthisloss}
\end{equation}

\section{Implementation}
To explore the large combination of PQCs previously mentioned, we introduce \textit{qregress} (Fig. \ref{fig:overview_project}), a modular Python package based on PennyLane\cite{bergholm_pennylane_2022} and Qiskit\cite{javadi-abhari_quantum_2024}.
The portion of the code based on PennyLane is capable of performing state-vector simulation using Qulacs\cite{suzuki_qulacs_2021}, noisy simulation using \textit{qiskit-aer} with the \textit{FakeQuebec} backend, along with access to real devices accessible on the IBM Quantum Platform using the PennyLane-Qiskit plugin.
Additionally, to train and test regression models that require many circuit executions, we ported our PennyLane code to Qiskit so that the Qiskit Batch Execution mode could be utilized.
This was a vital step for running on \textit{ibm\_quebec} since Qiskit Sessions, as implemented in the PennyLane-Qiskit plugin, are too cumbersome to train PQCs as $N$, the number of training samples, increases.
In practice, all state-vector simulations were performed using PennyLane, while all calculations using \textit{FakeQuebec} and \textit{ibm\_quebec}  are run using Qiskit.
All state-vector simulations with 5 qubits were optimized using 1000 iterations, while 16 qubits were initially run using 250 iterations.
State-vector simulations were also optimized using the Simultaneous Perturbation Stochastic Approximation method (SPSA), as implemented in PennyLane since we found that SPSA requires fewer steps to optimize when compared to other optimizers for state-vector simulation. 
For the experiments ran using \textit{FakeQuebec} and  \textit{ibm\_quebec}, we utilized the Constrained Optimization By Linear Approximation (COBYLA) optimizer, as implemented in SciPy\cite{virtanen_scipy_2020}.
For all models, features ($\mathbf{x}$) and target values ($\mathbf{y}$) were scaled using the MinMaxScaler, as implemented in Scikit-learn\cite{pedregosa_scikit-learn_2011}, such that all features and target values are $\mathbb{R}\in [ -1,1 ]$.

@@

An important aspect of ML in chemistry is the choice of molecular representation, or how the molecule is represented in the machine learning models.\cite{jones_molecular_2023}
In general, molecular representations can be partitioned into three groups, graph-, topology-, and physics-based representations, all of which were explored to provide a comprehensive overview of how they perform using classical models as a baseline.
During the preprocessing stage, before converting the XYZ coordinates into our molecular representation of choice, the set of hypothetical structures was reduced to 2436 molecules due to issues with valency exceptions when converting XYZ coordinates into RDKit mol objects.
Using RDKit\cite{noauthor_rdkit_nodate} we examined three commonly applied graph-based molecular preresentations, Molecular ACCess Systems (MACCS)\cite{durant_reoptimization_2002}, Morgan or extended-connectivity fingerprints \cite{morgan_generation_1965,rogers_extended-connectivity_2010}, and RDKit fingerprints.
All three of these methods use traversals of the molecular graphs to encode various structural details into bit vectors.
Lastly, we explore both topology- and physics-based molecular representations, both of which encode the three-dimensional structure of molecules in various, unique ways.
Persistent images (PIs) are a topology-based fingerprint that uses persistence homology to encode topological information of three-dimensional molecular structures into fixed-dimension images.\cite{adams_persistence_2017,townsend_representation_2020,schiff_augmenting_2022} 
We use the implementation from Townsend \textit{et al.}\cite{townsend_representation_2020}, which uses the Ripser Python package to generate PIs.\cite{tralie_ripserpy_2018}
Lastly, we explore two physics-based representations, Coulomb matrices (CMs) \cite{rupp_fast_2012} and smooth overlap of atomic positions (SOAPs), that were generated using DScribe.\cite{de_comparing_2016}
Both methods encode physical information regarding the atomic environments of each molecule, where CMs encode the Coulomb repulsion between atoms using nuclear charges, while SOAPs encode the similarities between atomic neighborhoods using kernels.
Due to the computational cost associated with computing the regularized entropy match (REMatch) kernel over the set of 2436 molecules with the SOAPs representation, we excluded SOAPs from the overall discussion.
For each of the molecular representations, we analyzed two different methods for representing the components of the BSE reaction.
The first represents the feature vector as the products subtracted from the reactants, denoted as \textit{sub} (e.g., $\mathbf{X}_{\mathit{sub}} = (\mathbf{X}_{\text{A}^{.}} + \mathbf{X}_{\text{B}^{.}}) - \mathbf{X}_{\text{A-B}}$), similar to the method used in Ref. \cite{garcia-andrade_barrier_2023}, and the second represents the feature vector using only the reactant (A-B), denoted as \textit{AB} (e.g., $\mathbf{X}_{\mathit{AB}} =  \mathbf{X}_{\text{A-B}}$).

Using this set of molecular representations, we also examine a diverse set of classical regression models to help choose the optimal representation for our diverse set of PQCs.
The classical models we analyze include ridge, lasso, elastic net, \textit{k}-nearest-neighbors, random forest, gradient boosting, support vector machines, kernel ridge, and Gaussian process regression as implemented in sckit-learn.\cite{pedregosa_scikit-learn_2011}
As highlighted in Fig. \ref{fig:classical_molrepfig}, we found that the best molecular representation across all models tested was Morgan fingerprints using the \textit{sub} formulation, which we use for all experiments using the PQCs.
Additionally, one challenge of applying classical molecular representations for ML tasks using PQCs is reducing the set of features, often containing hundreds or thousands of features per sample, to the number of qubits used on the quantum device.
For example, at the onset of this project, the standard number of qubits on the IBM quantum devices was 5 or 16 qubits, which requires reducing the initial set of 2048 features per sample for Morgan fingerprints down to 5 or 16 features.
To perform this reduction, we explored two different methods, SHapley Additive ExPlanation analysis (SHAP)\cite{lundberg_unified_2017}, which uses cooperative game theory to determine feature importance, and principal component analysis (PCA), which is a dimensionality reduction technique that captures the direction a datasets variance, as implemented in scikit-learn.\cite{pedregosa_scikit-learn_2011}
In Figs. \ref{fig:BSE_classical_features_R2} and \ref{fig:BSE_bse_classical_features_MAE}, we compare the two reduction techniques, for both 5 and 16 features, with the initial model containing 2048 features.
The initial model has mean absolute errors (MAE) of 1.91 and 4.98 kcal/mol and R$^{2}$s of 0.99 and 0.91 for the training and test set, respectively.
Using SHAP, we found that the model containing 5 features has an MAE of 16.08 kcal/mol and an R$^{2}$ of 0.39 for the training set, while the test set has an MAE of 15.86 kcal/mol and an R$^{2}$ of 0.42.
For the model with 16 features, we observe improvements in both the training and test sets, with MAEs of 10.48 and 11.08 kcal/mol and R$^{2}$ values of 0.69 and 0.68, respectively.
For the models using 5 and 16 features reduced through PCA, the training sets yield MAEs of 4.09 and 3.23 kcal/mol, while the test sets have MAEs of 10.17 and 8.40 kcal/mol, respectively.
The R$^{2}$s also show slight improvement over the feature reductions using SHAP, where the R$^{2}$s using 5 features are 0.95 and 0.69 for the training and test set, respectively, and the model using 16 features has R$^{2}$s of 0.97 and 0.78 for the training and test set, respectively.
Overall, the models using SHAP to reduce the feature set offer models with less overfitting, while sacrificing accuracy, when compared to the models reduced using PCA, which offer more accurate models that are more overfit. 
Based on these insights, for the experiments using PQCs, all BSE49 data is reduced to 5 or 16 qubits using PCA applied on Morgan fingerprint using the \textit{sub} formulation.

The second quantum chemistry dataset we analyze in this paper is generated using the data-driven coupled-cluster (DDCC) scheme of Townsend and Vogiatzis.\cite{townsend_data-driven_2019,jones_chapter_2023}
DDCC is an ML-based approach for accelerating the convergence of coupled-cluster singles and doubles (CCSD) calculations by predicting the $t_{2}$-amplitudes of the CCSD wave function (Eq. \ref{eq:cc_wfn}) with features generated using lower-level methods, such as Hartree-Fock (HF) and M{\o}ller-Plesset second-order perturbation theory (MP2), that are used to initialize CCSD calculations.
Herein, we provide a brief overview of the equations required to understand the DDCC method, starting with the coupled-cluster wave function, which takes the general form,
\begin{equation}
	\ket{\Psi_{\text{CC}}} = \exp(\hat{T}) \ket{\Psi_{0}}
	\label{eq:cc_wfn}
\end{equation}
where $\hat{T}$ is the cluster operator and $\ket{\Psi_{0}}$ is the reference (Hartree-Fock) wave function.
In CCSD, the $\hat{T}$ operator is truncated to only include single ($\hat{T}_{1}$) and double ($\hat{T}_{2}$) excitations.
After solving the coupled-cluster equations, the CCSD correlation energy can be formulated as,
\begin{equation}
	E^{\text{CCSD}}_{\text{corr}} = \sum_{\substack{a<b \\ i<j}} \mel{ij}{}{ab} t^{ab}_{ij} + \sum_{\substack{a<b \\ i<j}} \mel{ij}{}{ab} t^{a}_{i} t^{b}_{j}
	\label{eq:cc_corr}
\end{equation}
where $i$ and $j$ denote occupied orbitals, $a$ and $b$ denote virtual orbitals, $t^{ab}_{ij}$ correspond to two-electron excitation amplitudes ($t_{2}$-amplitudes), $t^{a}_{i}$ and $t^{b}_{j}$ correspond to one-electron excitation amplitudes ($t_{1}$-amplitudes), and $\mel{ij}{}{ab}$ are two-electron integrals.
The objective of the DDCC method is to learn the CCSD $t_{2}$-amplitudes using features generated using HF and MP2, since the CCSD $t_{2}$-amplitudes are initialized using MP2 $t_{2}$-amplitudes, defined as,
\begin{equation}
	t^{ab}_{ij(\text{MP2})} = \frac{\mel{ij}{}{ab}}{\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}}
	\label{eq:MP2_t2}
\end{equation}
where $\varepsilon_{i}$ and $\varepsilon_{j}$ denote the orbital energies of the occupied orbitals $i$ and $j$, while the virtual orbitals $a$ and $b$ are denoted by $\varepsilon_{a}$ and $\varepsilon_{b}$.
Due to this important relationship, MP2 $t_{2}$-amplitudes are included in the DDCC feature set, along with the numerator ($\mel{ij}{}{ab}$), denominator ($\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}$), a binary feature to denote whether the excitation goes to the same virtual orbital, and the orbital energies ($\varepsilon_{i},\varepsilon_{j},\varepsilon_{a},\varepsilon_{b}$).
Additionally, the feature set includes terms related to the individual contributions to the orbital energies, such as the one-electron Hamiltonian ($h$), Coulomb matrix ($J$), exchange matrix $K$, and Coulomb and exchange integrals ($J^{i}_{a}, J^{j}_{b}, K^{a}_{i}, K^{b}_{j}$).
In total, there are 30 features for each CCSD $t_{2}$-amplitude due to the addition of features that denote the sign and magnitude of the MP2 $t_{2}$-amplitudes.
